FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories

### 2025-01-14 manuscript.Rmd
Update manuscript.Rmd
Co-authored-by: Maximilian Ernst <34346372+Maximilian-Stefan-Ernst@users.noreply.github.com>

A third impediment arises when theories have a "dependency on the author" (DOA). <!-- AB citation needed?! Who came up with this term? -->
DOA occurs when a theory cannot be understood by independent scholars,
thus requiring the original author for interpretation and clarification.
We have heard DOA referred to apocryphally as the "ask Leon" phenomenon,
as graduate students were supposedly told to ask Leon Festinger to explain to them how their misconstrual of cognitive dissonance theory had caused their experiments to yield null results.
DOA relates to the discourse on "Great Man Theorizing" [@guestWhatMakesGood2024] because it enables gatekeeping: an author could insist that work requires their involvement or denounce work conducted outside their purview as illegitimate,
which violates checks and balances of scientific research.
DOA also renders theories immune to refutation,
because the author can claim that the theory was misconstrued when confronted with falsifying evidence, thus making it a moving target  [@szollosiArrestedTheoryDevelopment2021].
The fact that DOA is inherently problematic is illustrated by cases where third parties identify logical inconsistencies within a theory [e.g., @kissnerIDENTIFICATIONLOGICALINCONSISTENCY2008].
This demonstrates that original authors are not the ultimate authority on their theories.
DOA thus unduly impedes scientific progress, and authors should make good-faith efforts to make theories as accessible as possible; both in terms of availability and in terms of interpretability.

>>>>

A third impediment arises when theories have a "dependency on the author" (DOA). <!-- AB citation needed?! Who came up with this term? -->
DOA occurs when a theory cannot be understood by independent scholars,
thus requiring the original author for interpretation and clarification.
We have heard DOA referred to apocryphally as the "ask Leon" phenomenon,
as graduate students were supposedly told to ask Leon Festinger to explain to them how their misconstrual of cognitive dissonance theory had caused their experiments to yield null results.
DOA relates to the discourse on "Great Man Theorizing" [@guestWhatMakesGood2024] because it enables gatekeeping: an author could insist that work requires their involvement or denounce work conducted outside their purview as illegitimate,
which violates checks and balances of scientific research.
DOA also renders theories immune to refutation,
because the author can claim that the theory was misconstrued when confronted with falsifying evidence, thus making it a moving target  [@szollosiArrestedTheoryDevelopment2021].
The fact that DOA is inherently problematic is illustrated by cases where third parties identify logical inconsistencies within a theory [e.g., @kissnerIDENTIFICATIONLOGICALINCONSISTENCY2008].
This demonstrates that original authors are not the ultimate authority on their theories.
DOA thus unduly impedes scientific progress, and authors should make good-faith efforts to make theories as accessible as possible in terms of both availability and interpretability.
